Typed Lua: An Optional Type System for Lua

### 2014-03-06 doc/paper/typedlua.tex
rewrote section on union types

Typed Lua includes union types as a way to discriminate data
structures that can hold values of several different, but fixed types.
Next, we show an example:
\begin{verbatim}
    local x:number?            -- OK
    local y:number|string = 1  -- OK
    local z:number = x         -- Error
\end{verbatim}

The annotation \texttt{number?} is syntactic sugar for
\texttt{number|nil} that means the type $\Number\;\cup\;\Nil$.
Typed Lua includes the syntactic sugar \texttt{t?} because the
type $t\;\cup\;\Nil$ is very common in Lua code.

The last line of the previous example gives a type error because we
are attempting to assign a value of type $\Number\;\cup\;\Nil$
to a variable that accepts only values of type $\Number$.
Although we can solve this type error changing the annotation of
variable \texttt{z} from \texttt{number} to \texttt{number?}, there is
another way that we can use to solve this type error without changing
the annotation:
\begin{verbatim}
    local z:number = x or 0  -- OK
\end{verbatim}

Typed Lua has a typing rule for checking that
$t\;\cup\;\Nil\;\mathbf{or}\;t$ always has the type $t$.

>>>>

Typed Lua uses union types to encode some common Lua idioms:
optional values, overloading based on the tags of input parameters,
and overloading on the return type of the functions.

Optional values are unions of some type and $\Nil$, and are so
common that Typed Lua uses the {\tt t?} syntax for these unions.
They appear any time a function has optional parameters, and
any time the program reads a value from an array or map.

\begin{verbatim}
    local function message(name: string,
                           greeting: string?)
      local greeting = greeting or "Hello "
      return greeting .. name
    end

    print(message("Lua"))
    print(message("Lua", "Hi"))
\end{verbatim}

In this example, the second parameter is optional but, in the first
line of the function, we declare a new variable that is guaranteed to
have type $\String$ instead of $\String \cup \Nil$. In Lua, any value
except {\tt nil} and {\tt false} are "truthy", so the short-circuiting
{\tt or} operator is a common way of giving a default value to an
optional parameter. Typed Lua encodes this idiom with a typing rule: if
the left side of {\tt or} has type $T \cup Nil$ and the right side
has type $T$ then the {\tt or} expression has type $T$.

Declaring a new {\tt greeting} variable that shadows the parameter
is not necessary:

\begin{verbatim}
    local function message(name: string, 
                           greeting: string?)
      greeting = greeting or "Hello "
      return greeting .. name
    end
\end{verbatim}

Typed Lua lets assignment change the type of a local variable in
cases that respect a simple heuristic: the new type must be a subtype
of the previous one, and only applies for the remainder of the
current scope.

Overloaded functions use the {\tt type} function to inspect
the tag of their parameters, and perform different actions depending
on what those tags are. The simplest case overloads on just a
single parameter:

\begin{verbatim}
    local function overload(s1: string, 
                            s2: string|number)
      if type(s2) == "string" then
        return s1 .. s2
      else
        -- string.rep: (string, number) -> string
        return string.rep(s1, s2)
      end
    end
\end{verbatim}

Typed Lua has a small set of type predicates that, when used
over a local variable in a condition, constrain the type of that
variable. The function above uses the {\tt type(X) == "string"}
predicate which constrains the type of $X$ from $T \cup \String$ to
$\String$ when the predicate is true and $T$ otherwise. This is
a simplified form of {\em flow typing}~\cite{guha:flow}.

The type predicates can only discriminate based on tags, so they
are limited on the kinds of unions that they can discriminate. It
is possible to discriminate a union that combines a table type with
a base type, or a table type with a function type, or a two base types,
but it is not possible to discriminate between two different function
types.

Functions that overload their return types to signal the ocurrence
of errors are another common Lua idiom. In this idiom, a function
returns its normal set of return values in case of success but,
if anything fails, returns {\tt nil} as the first value, followed
by an error message or other data describing the error, as in the
following example:

\begin{verbatim}
    local function idiv(d1: number, d2: number):
          (number, number)|(nil, string)
      if d2 == 0 then
        return nil, "division by zero"
      else
        local r = d1 % d2
        local q = (d1 - r)/d2
        return q, r
      end
    end
\end{verbatim}

There is also special syntax for this idiom: we could
annotate the return type of {\tt idiv} with {\tt (number, number)?}
to denote the same union\footnote{The parentheses are always
necessary here: {\tt number?} is {\tt number|nil}, while {\tt (number)}
is {\tt (number)|(nil, string)}.}.

The full type of {\tt idiv} is $\Number \times \Number \times
\Top * \rightarrow (\Number \times \Number \times \Nil *) \cup
(\Nil \times \String \times \Nil *)$. A typical client of this
function would use it as follows:

\begin{verbatim}
    local q, r = idiv(n1, n2)
    -- q is number|nil, r is number|string
    if q then
      -- q and r are numbers
    else
      -- r is a string
    end
\end{verbatim}

When Typed Lua encounters a union of tuples in the right side
of an declaration, it stores the the union in a special type
environment with a fresh name and assigns {\em projection types} 
to the variables in the left side of the declaration. If the type
variable is $X$, variable {\tt q} gets type $X_1$ and variable
{\tt r} gets type $X_2$.

If we need to check a projection type
against some other type, we take the union of the corresponding
component in each tuple. But if we a variable with a projection
type appears in a type predicate, the predicate discriminates
against all tuples in the union. In the example above,
$X$ is $(\Number \times \Number \times \Nil *) \cup
(\Nil \times \String \times \Nil *)$ outside of the {\tt if}
statement, but $\Number \times \Number \times \Nil *$ in the
{\tt then} block and $\Nil \times \String \times \Nil *$ in
the {\tt else} block.

Notice that we could also discriminate using {\tt r}, using
{\tt type(r) == "number"} as our predicate, with the same
result. The first form is more succint, and more idiomatic.
We can also use projection types to write overloaded functions
where the type of a parameter depends on the type of another
parameter.

Currently, a limitation of our overloading mechanisms is that
the return type cannot depend on the input types; we cannot
write a function that is guaranteed to return a number if
passed a number and guaranteed to return a string if passed
a string, for example.
